Inferring cosmology from gravitational waves using non-parametric detector-frame mass distribution

### 2024-07-18 src/tex/ms.tex
[showyourwork] overleaf sync

\begin{table}
    \label{tab:notation}
    \caption{Notation}
    \begin{ruledtabular}
        \begin{tabular}{c>{\raggedright\arraybackslash}p{0.8\linewidth}}
            Notation & Description \\
            \hline
            $m_1$ & Primary mass of the \ac{GW} sources \\
            $z$ & Redshift of the \ac{GW} sources \\
            $d_L$ & Luminosity distance of the \ac{GW} sources \\
            $m^z_1$ & Primary mass of the \ac{BBH} sources in the detector frame \\
            $Y_t$ & source-frame primary mass posterior samples of the t-th \ac{GW} event \\
            $\mathbf{Y}$ & Set of source-frame primary mass posterior samples of all \ac{GW} events \\
            $\mathbf{\Theta}$ & Set of hyperparameters of the \ac{(H)DPGMM} \\
            $p(m^z_1|\mathbf{\Theta})$ & Detector-frame observed mass population distribution reconstructed with \ac{(H)DPGMM} \\
            $\Lambda$ & Source-frame population model \\
            $\Omega$ & Cosmological model \\
            $p(m_1|\Lambda)$ & Source-frame mass population distribution assuming the population model $\Lambda$ \\
            $p(m^z_1|\Lambda, \Omega)$ & Detector-frame mass population distribution assuming the population model $\Lambda$ and cosmological model $\Omega$ \\
            $\mathrm{det}$ & detectability of the \ac{GW} events \\
        \end{tabular}
        \end{ruledtabular}
\end{table}

@@

On the other hand, we assume a population model $\Lambda$ to obtain a source-frame mass population $p(m_1|\Lambda)$, which represents the underlying population of the sources.
Note that, either a parametric model can be used to only assume the form of the population distribution, or a fixed population model can be used to assume an exact population distribution.
We then transform $p(m_1|\Lambda)$ to the detector frame with a given cosmological model $\Omega$ to obtain the detector-frame mass population $p(m^z_1|\Lambda, \Omega)$ by
\begin{widetext}
    \begin{equation}
        \label{eq:transformation}
        \begin{split}
            &p(m^z_1|\Lambda, \Omega, \mathrm{det}) \\
            &= \int p(m^z_1|m_1, z, \Lambda, \Omega, \mathrm{det}) p(m_1, z|\Lambda, \Omega, \mathrm{det}) \mathrm{d}m_1 \mathrm{d}z \\
            &= \int p(m^z_1|m_1, z) \frac{p(\mathrm{det}|m_1, z, \Lambda, \Omega)p(m_1, z|\Lambda, \Omega)}{p(\mathrm{det}|\Lambda, \Omega)} \mathrm{d}m_1 \mathrm{d}z \\
            &= \frac{1}{p(\mathrm{det}|\Lambda, \Omega)}\int \delta(m^z_1-m_1(1+z))p(\mathrm{det}|m_1, z, \Omega)p(m_1|\Lambda)p(z|\Lambda) \mathrm{d}m_1 \mathrm{d}z \\
            &\propto \int p\left(\mathrm{det}\middle|\frac{m^z_1}{1+z},z,\Omega\right)p\left(\frac{m^z_1}{1+z}\middle|\Lambda\right)p(z|\Lambda) \mathrm{d}z\,,
        \end{split}
    \end{equation}
\end{widetext}
where we assume the source-frame mass distribution does not evolve with redshift, and $\mathrm{det}$ represents whether the event is detectable or not.
The selection effect is included in the derivation in eq.~\eqref{eq:transformation}, specifically, the first term in the final expression, which is the selection function.
Note that the selection function in general is a function of the detector-frame mass $m^z_1$, luminosity distance $d_L$, and mass ratio $q$.
In this work, we marginalized over $q$ to study the primary mass distribution only.
Due to the censoring of the low and high mass events, we consider the selection function in the transformation instead of the reconstruction.

@@

For each reconstructed distribution $p(m^z_1|\Theta_i)$, we find the optimized $\Lambda_j$ and $\Omega_j$ by minimizing $d_\mathrm{JS}$.
For the minimization, we use the \textsc{scipy} \citep{2020SciPy-NMeth} implementation of the \todo{XXX} algorithm. \todo{Add reference for the algorithm}
This procedure gives us a set of optimized $\Lambda_i$ and $\Omega_j$ for each $\Theta_i$.
Note that we first calculate the median of the detector-frame primary mass posterior samples of each \ac{GW} event, then we calculate $d_\mathrm{JS}$ in the range from the minimum to the maximum of the medians.
This is because the reconstruction is only valid in the region where observations are present.
In Fig.~\ref{fig:simulation_comparison}, we show an illustration of the comparison between the two populations.

The resulting distribution of the optimized parameters is not a posterior distribution in the Bayesian sense, as we do not consider the likelihood of the observed data given the population model.
Instead, it is a distribution of the optimized parameters that are consistent with the observed data, distributed according to the uncertainty in the reconstruction of the population.
That is, the uncertainty in the comparison of the two populations is not considered in the resulting distribution.
By considering this optimization procedure only, we can estimate the optimized parameters without performing computationally expensive likelihood calculations.

The analysis code is publicly available.\footnote{\todo{Add link}}

@@

>>>>

\begin{table}
    \label{tab:notation}
    \caption{Notation}
    \begin{ruledtabular}
        \begin{tabular}{c>{\raggedright\arraybackslash}p{0.8\linewidth}}
            Notation & Description \\
            \hline
            $m_1$ & Primary mass of the \ac{GW} sources \\
            $q$ & Mass ratio of the \ac{GW} sources \\
            $z$ & Redshift of the \ac{GW} sources \\
            $d_L$ & Luminosity distance of the \ac{GW} sources \\
            $m^z_1$ & Primary mass of the \ac{BBH} sources in the detector frame \\
            $Y_t$ & source-frame primary mass posterior samples of the $t$-th \ac{GW} event \\
            $\mathbf{Y}$ & Set of source-frame primary mass posterior samples of all \ac{GW} events \\
            $\mathbf{\Theta}$ & Set of hyperparameters of the \ac{(H)DPGMM} \\
            $p(m^z_1|\mathbf{\Theta})$ & Detector-frame observed mass population distribution reconstructed with \ac{(H)DPGMM} \\
            $\Lambda$ & Source-frame population model \\
            $\Omega$ & Cosmological model \\
            $p(m_1|\Lambda)$ & Source-frame mass population distribution assuming the population model $\Lambda$ \\
            $p(m^z_1|\Lambda, \Omega)$ & Detector-frame mass population distribution assuming the population model $\Lambda$ and cosmological model $\Omega$ \\
            $\mathrm{det}$ & detectability of the \ac{GW} events \\
        \end{tabular}
        \end{ruledtabular}
\end{table}

@@

On the other hand, we assume a population model $\Lambda$ to obtain a source-frame mass population $p(m_1|\Lambda)$, which represents the underlying population of the sources.
Note that, either a parametric model can be used to only assume the form of the population distribution, or a fixed population model can be used to assume an exact population distribution.
We then transform $p(m_1|\Lambda)$ to the detector frame with a given cosmological model $\Omega$ to obtain the detector-frame mass population $p(m^z_1|\Lambda, \Omega)$ by
\begin{widetext}
\begin{equation}
    \label{eq:transformation}
    \begin{aligned}
        &p(m^z_1|\Lambda, \Omega, \mathrm{det}) \\
        &= \int p(m^z_1|m_1, z, \Lambda, \Omega, \mathrm{det}) p(m_1, z|\Lambda, \Omega, \mathrm{det}) \mathrm{d}m_1 \mathrm{d}z \\
        &= \int p(m^z_1|m_1, z) \frac{p(\mathrm{det}|m_1, z, \Lambda, \Omega)p(m_1, z|\Lambda, \Omega)}{p(\mathrm{det}|\Lambda, \Omega)} \mathrm{d}m_1 \mathrm{d}z \\
        &= \frac{1}{p(\mathrm{det}|\Lambda, \Omega)}\int \delta(m^z_1-m_1(1+z))p(\mathrm{det}|m_1, z, \Omega)p(m_1|\Lambda)p(z|\Lambda) \mathrm{d}m_1 \mathrm{d}z \\
        &\propto \int p\left(\mathrm{det}\middle|\frac{m^z_1}{1+z},z,\Omega\right)p\left(\frac{m^z_1}{1+z}\middle|\Lambda\right)p(z|\Lambda) \mathrm{d}z\,,
    \end{aligned}
\end{equation}
\end{widetext}
where we assume the $m_1$ distribution does not evolve with $z$, and $\mathrm{det}$ represents whether the event is detectable or not.
The selection effect is included in the derivation in eq.~\eqref{eq:transformation}, specifically, the first term in the final expression, which is the selection function.
Note that the selection function in general is a function of the detector-frame mass $m^z_1$, luminosity distance $d_L$, and mass ratio $q$.
In this work, we marginalized over $q$ to study the primary mass distribution only.
Due to the censoring of the low and high mass events, we consider the selection function in the transformation instead of the reconstruction.

@@

For each reconstructed distribution $p(m^z_1|\Theta_i)$, we find the optimized $\Lambda_j$ and $\Omega_j$ by minimizing $d_\mathrm{JS}$.
For the minimization, we use the \textsc{scipy} \citep{2020SciPy-NMeth} implementation of the \todo{XXX} algorithm. \todo{Add reference for the algorithm}
This procedure gives us a set of optimized $\Lambda_i$ and $\Omega_j$ for each $\Theta_i$.
Note that we calculate $d_\mathrm{JS}$ in the range from the minimum to the maximum of the median of the $m^z_1$ posterior samples of each \ac{GW} event
This is because the reconstruction is only valid in the region where observations are present.
In Fig.~\ref{fig:simulation_comparison}, we show an illustration of the comparison between the two populations.

The resulting distribution of the optimized parameters is not a posterior distribution in a Bayesian sense, as we do not consider the likelihood of the observed data given the population model.
Instead, it is a distribution of the optimized parameters that are consistent with the observed data, distributed according to the uncertainty in the reconstruction of the population.
That is, the uncertainty in the comparison of the two populations is not considered in the resulting distribution.
By considering this optimization procedure only, we can estimate the optimized parameters without performing computationally expensive likelihood calculations.

The analysis code is publicly available.\footnote{The analysis code is available at \url{https://github.com/thomasckng/det-frame-cosmo-with-FIGARO/tree/main/src/scripts}.}

@@

In this section, we test our method with a mock data study.
To prepare the mock data, we first generated $10000$ sets of $m_1$ and $z$ samples from a \textsc{power law + peak} model for the $m_1$ distribution and a uniform distribution in comoving volume for the $z$ distribution.
Here we assumed the $m_1$ distribution does not evolve with $z$.
we then transformed the $m_1$ and $z$ samples to $m^z_1$ and $d_L$ samples.
After the transformation, we applied a selection function to the $m^z_1$ and $d_L$ samples.
We chose the O3 selection function from \todo{add reference} as the selection function.
Note that we marginalized over $q$ for the selection function, similar to the transformation in Sec.~\ref{sec:transformation}.
After applying the selection function, we generated the $m^z_1$ posterior samples for each true $m^z_1$ sample by
\begin{gather*}
    I \sim \mathcal{N}(0, 0.03)\,,\\
    P \sim \mathcal{N}(\log(T_t)+I, 0.03)\,,\\
    Y_t = \exp(P)\,,
\end{gather*}
where $\mathcal{N}(\mu, \sigma)$ is a normal distribution with mean $\mu$ and standard deviation $\sigma$, $I$ is the intrinsic error, $P$ is the posterior sample in the log space, $T_t$ is the true $m^z_1$ sample, and $Y_t$ is the posterior sample for the $t$-th \ac{GW} event.
